---
title: "Data Cleaning and Descriptive Statistics"
author: "Allison McCarty"
date: "2/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Install Libraries
library(readxl)
library(tidyverse)
library(dplyr)
library(pastecs)
library(data.table)
library(formattable)
library(tidyr)

dat1 <- read_excel("Practicum 1 Data.xlsx")
sum(apply(dat1, 1, function(x){any(is.na(x))})) # every single observation has a N/A for at least one variable
```

The first obvious challenge is that this dataset does not include a single complete case.  In otherwords, every single observation at least has one variable populated with "N/A".  Thus, using a function such as complete.cases would be highly problematic in this circumstance (we would have no data left). 

There are also a lot of variables.  Let's focus specifically on the 13 mentioned in the Write-Up.

```{r datacleaning}
#Select variables of interest
myvars <- c("shift", "gender", "Age", "height", "weightgain", "lbs_gained", "pounds_gained", "bweight", "BMI", "VigTime", "WalkTime", "ModTime", "Total_Met_Min")
pracdat <- dat1[myvars]

#Check for observations with any NAs in the filtered dataset
sum(apply(pracdat, 1, function(x){any(is.na(x))})) #Most (264/392) have some N/As, therefore we still cannot do a full N/A omit for this data

#Get rid of observations that do not have ANY variables populated
all.na <- (apply(pracdat, 1, function(x){all(is.na(x))}))
pracdat <- data.frame(pracdat, all.na)
pracdat <- subset(pracdat, all.na == "FALSE")

# Go back and overwrite Total metabolic minutes
# Here is the equation to find total metabolic minutes from the other data available to us
# Total Metabolic Mintues = 8*VigTime + 4*ModTime + 3.3*WalkTime
# Perform the calculation for all of the observations and overwrite the column in the original dataframe
pracdat$Total_Met_Min = 8*pracdat$VigTime + 4*pracdat$ModTime + 3.3*pracdat$WalkTime
sum(apply(pracdat, 1, function(x){any(is.na(x))})) # Now we have 189 complete cases (61 more than before)

# Remove observations that do not have data for total metabolic minutes, shift, lbs gained, or weight gain (binary)
pracdat <- pracdat[complete.cases(pracdat[c(1, 5, 6, 13)]),]
```

Now, lets do descriptive statistics
```{r num_datacleaning}
#Only for qualitative
summary(pracdat[3:13])


quandat <- stat.desc(pracdat[3:13])
options(scipen=100)
options(digits=2)
descriptives.df <- stat.desc(pracdat[c(3:4, 7:13)], basic = F)
names(descriptives.df) <- c("Age", "Height", "Pounds Gaines", "Body Weight", "BMI", "Vigorous Activity Minutes", "Walking Minutes", "Moderate Activity Minutes", "Total Metabolic Minutes")
formattable(descriptives.df, align =c("l", rep("c", rep(8)), "r"))


#Let's also do the categorical
#Gender First
(gender.df <- table(pracdat$gender))
gender.df<- data.frame(gender.df)
colnames(gender.df)[1] <- "Gender"


#Now Shift
(shift.df <- table(pracdat$shift))
shift.df<- data.frame(shift.df)
colnames(shift.df)[1] <- "Shift"
formattable(shift.df)

#Visualize the categorical variables (probably will not be included in practicum 1)
library(ggpubr)
ggbarplot(gender.df, x = "Gender", y = "Freq", combine = TRUE)
ggbarplot(shift.df, x = "Shift", y = "Freq", fill= "cornflowerblue", main = "Shift distribution of Sample Population" )

ggplot(pracdat, aes(x = shift)) + geom_dotplot(binwidth = .1)
```
