---
title: "Praticum 1 Technical Report"
author: "Allison McCarty"
date: "2/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Install Libraries
library(readxl)
library(tidyverse)
library(dplyr)
library(pastecs)
library(data.table)
library(formattable)
library(tidyr)
library(kableExtra)
library(gridExtra)
library(ggpubr)
library(ggthemes)
```

```{r InitialDataCleaning}
#Import Data
dat1 <- read_excel("Practicum 1 Data.xlsx")

#From the SAP, we decided that we would include 13 variables in the analysis for this project
#Select variables of interest
myvars <- c("shift", "gender", "Age", "height", "weightgain", "lbs_gained", "pounds_gained", "bweight", "BMI", "VigTime", "WalkTime", "ModTime", "Total_Met_Min")
pracdat <- dat1[myvars]

#Get rid of observations that do not have ANY variables populated
all.na <- (apply(pracdat, 1, function(x){all(is.na(x))}))
pracdat <- data.frame(pracdat, all.na)
pracdat <- subset(pracdat, all.na == "FALSE")

#Get rid of the last column that we used to subset complete cases
pracdat <- pracdat[1:13]

# Next, we will go back and manually calculate metabolic minutes and overwrite the column in the original data frame
# Calculation for Total Met Minutes = 8*VigTime + 4*ModTime + 3.3*WalkTime
pracdat$Total_Met_Min = 8*pracdat$VigTime + 4*pracdat$ModTime + 3.3*pracdat$WalkTime
sum(apply(pracdat, 1, function(x){any(is.na(x))})) # Now we have 189 complete cases (61 more than before)

#Next, we will go back and manually calculate BMI and overwite the column in the original data frame
# ***INCLUDE JOE'S CODE HERE***

#Now, let's see what variables are missing for each of the predictors
sum(apply(pracdat, 1, function(x){any(is.na(x))})) # Now we have 189 complete cases (61 more than before)

sum(complete.cases(pracdat[1])) #348 out of 352 observations have data on shift
sum(complete.cases(pracdat[5])) #348 out of 352 observations have data on binary weight gain
sum(complete.cases(pracdat[13])) #351 of the 352 observations have metabolic minutes data (meaning they also have data on walk, vig, and mod ex time)
sum(complete.cases(pracdat[6])) #232 of the 352 observations have data for lbs gained
```

Commentary on lack of data for lbs gained: 

There are a lot of N/A values for lbs gained because we will not have data for a specific observation unless they gained weight. If a person lost weight in the study, this would be populated as "N/A" for lbs gained (col 6). For the purposes of the second model, we may choose to select the subset of the population who gained weight (have data for this variable)/or did not gain weight (do NOT have data for this variable) and create a model specific to that subset


For Model 1, we will just use the binary weight gain variable to create a logistic model.  For Model 2, we will subset the sample to those who gained weight, and build multiple models to determine which factors influenced this weight gain.

```{r MoreDataCleaning}
#Create final data frame for Model #1
pracdat.mod1 <- pracdat[-6:-7]
#Change Weight Gain Factor Variable to Binary
pracdat.mod1$weightgain <- ifelse(pracdat.mod1$weightgain=="Yes",1,0)


# Create subset data frames for Model #2
pracdat.mod2 <- subset(pracdat, weightgain == "Yes")
```
