---
title: "Praticum 1 Technical Report"
author: "Allison McCarty"
date: "2/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Install Libraries
library(readxl)
library(tidyverse)
library(dplyr)
library(pastecs)
library(data.table)
library(formattable)
library(tidyr)
library(kableExtra)
library(gridExtra)
library(ggpubr)
library(ggthemes)
library(mgcv)
```

```{r InitialDataCleaning, include=FALSE}
#Import Data
dat1 <- read_excel("Practicum 1 Data.xlsx")

#From the SAP, we decided that we would include 13 variables in the analysis for this project
#Select variables of interest
myvars <- c("shift", "gender", "Age", "height", "weightgain", "lbs_gained", "pounds_gained", "bweight", "BMI", "VigTime", "WalkTime", "ModTime", "Total_Met_Min")
pracdat <- dat1[myvars]

#Get rid of observations that do not have ANY variables populated
all.na <- (apply(pracdat, 1, function(x){all(is.na(x))}))
pracdat <- data.frame(pracdat, all.na)
pracdat <- subset(pracdat, all.na == "FALSE")

#Get rid of the last column that we used to subset complete cases
pracdat <- pracdat[1:13]

# Next, we will go back and manually calculate metabolic minutes and overwrite the column in the original data frame
# Calculation for Total Met Minutes = 8*VigTime + 4*ModTime + 3.3*WalkTime
pracdat$Total_Met_Min = 8*pracdat$VigTime + 4*pracdat$ModTime + 3.3*pracdat$WalkTime

#Next, we will go back and manually calculate BMI and overwite the column in the original data frame

pracdat$Begin_bweight = ifelse(is.na(pracdat$lbs_gained), pracdat$bweight, (pracdat$bweight)-(pracdat$lbs_gained))
pracdat$Before_BMI = round((703*pracdat$Begin_bweight)/(pracdat$height*pracdat$height),2)
pracdat$BMI_change = (pracdat$BMI)-(pracdat$Before_BMI)

#Now, let's see what variables are missing for each of the predictors
sum(apply(pracdat, 1, function(x){any(is.na(x))})) # Now we have 189 complete cases (61 more than before)

sum(complete.cases(pracdat[1])) #348 out of 352 observations have data on shift
sum(complete.cases(pracdat[5])) #348 out of 352 observations have data on binary weight gain
sum(complete.cases(pracdat[13])) #351 of the 352 observations have metabolic minutes data (meaning they also have data on walk, vig, and mod ex time)
sum(complete.cases(pracdat[6])) #232 of the 352 observations have data for lbs gained
```


### Exploratory Data Analysis:

Visuals for Categorical Variables

```{r}
ggplot(pracdat, aes(x=gender, y=BMI_change)) + geom_boxplot()
ggplot(pracdat, aes(x=Age, y = BMI)) + geom_point(aes(color=gender))

#solid
ggplot(na.omit(pracdat), aes(x = Age, y = Total_Met_Min, color = gender)) + geom_line(size = .8) + facet_grid(. ~ gender) + theme_bw() + theme(panel.grid = element_blank())


ggplot(pracdat, aes(x= gender, y =BMI_change)) + geom_boxplot(aes(color = gender)) + facet_grid(. ~ shift)

#solid
ggplot(na.omit(pracdat), aes(x= gender, y =BMI_change)) + geom_boxplot(aes(color = gender)) + facet_grid(. ~ shift)

#decent
p1 <- ggplot(na.omit(pracdat), aes(x = shift, y = VigTime, fill=gender)) + geom_bar(stat="identity") + theme_minimal() + theme(axis.text.x = element_text(angle=90, hjust=1))
p2 <- ggplot(na.omit(pracdat), aes(x = shift, y = WalkTime, fill=gender)) + geom_bar(stat="identity") + theme_minimal() +theme(axis.text.x = element_text(angle=90, hjust=1))
p3 <- ggplot(na.omit(pracdat), aes(x = shift, y = ModTime, fill=gender)) + geom_bar(stat="identity") + theme_minimal() + theme(axis.text.x = element_text(angle=90, hjust=1))
p4 <- ggplot(na.omit(pracdat), aes(x = shift, y = Total_Met_Min, fill=gender)) + geom_bar(stat="identity") + theme_minimal() + theme(axis.text.x = element_text(angle=90, hjust=1))
grid.arrange(p1, p2, p3, p4, nrow = 2)


```


Test for normality of the continuous variables:

```{r}
#First, we want to make qqnorm plots for all of the continuous variables, which will be used to justify transformations and included in the appendix
library(car)
pracdat.cont <- pracdat[c(-1:-2, -5:-6)]
cont.labels <- colnames(pracdat.cont)
newfun <- function(x, y) qqPlot(x, main = paste(y, "QQ Plot", sep = " "))
par(mfrow = c(1, 2))
mapply(newfun, pracdat.cont, cont.labels)
```

We made QQPlots for all of the continuous numeric variables.  Based on the plots, we think that age, height, body weight, and BMI are approximately normal (at least normal enough that they do not merit transformations).

Pounds gained, VigTime, ModTime, and WalkTime, and TotalMetMinutes do not follow a normal distribution and have a strong right skew. We will take a closer look at these variables and try a log transformation if necessary.

```{r}
pracdat.abnormal <- pracdat.cont[c(3, 6:9)]
library(reshape2)
pracdat.abnormal<- melt(pracdat.abnormal)
ggplot(data = pracdat.abnormal , aes(x = value)) + stat_density() + facet_wrap(~variable, scales = "free")
#Most have an extreme right skew

#Now, let's take a log transformation of all of these variables
ggplot(data = pracdat.abnormal , aes(x = log(value))) + stat_density() + facet_wrap(~variable, scales = "free")
# Much better


#Perform the transformation to overall dataframe
pracdat$lbs_gained <- log(pracdat$lbs_gained+1)
pracdat$pounds_gained <- log(pracdat$pounds_gained+1)
pracdat$VigTime <- log(pracdat$VigTime+1)
pracdat$WalkTime <- log(pracdat$WalkTime+1)
pracdat$ModTime <- log(pracdat$ModTime+1)
pracdat$Total_Met_Min <- log(pracdat$Total_Met_Min+1)
```

**NOTE THAT ANY DATAFRAME AFTER THIS LINE DERIVED FROM PRACDAT HAS LOG TRANSFORMATIONS IN POUNDS GAINED AND EXERCISE MINUTES ESTIMATES**

### Model Analysis
Commentary on lack of data for lbs gained and motivation for building two separate models: 

There are a lot of N/A values for lbs gained because we will not have data for a specific observation unless they gained weight. If a person lost weight in the study, this would be populated as "N/A" for lbs gained (col 6). For the purposes of the second model, we may choose to select the subset of the population who gained weight (have data for this variable)/or did not gain weight (do NOT have data for this variable) and create a model specific to that subset.


For Model 1, we will just use the binary weight gain variable to create a logistic model.  For Model 2, we will subset the sample to those who gained weight, and build multiple models to determine which factors influenced this weight gain.

Data Cleaning specific to our models:
```{r MoreDataCleaning}
#Create final data frame for Model #1
pracdat.mod1 <- pracdat[-6:-7]
#Change Weight Gain Factor Variable to Binary
pracdat.mod1$weightgain <- ifelse(pracdat.mod1$weightgain=="Yes",1,0)


# Create subset data frames for Model #2
pracdat.mod2 <- subset(pracdat, weightgain == "Yes")
```

Binary WeightGain Model **should use pracdat.mod1**

```{r}
#weightglm1 = glm(weightgain ~ VigTime + WalkTime + ModTime, data = pracdat, family = "binomial)
#weightglm2 = glm(weightgain ~  , family = "binomial)
#weightglm3 = glm(weightgain ~  , family = "binomial)
#weightglm4 = glm(weightgain ~  , family = "binomial)

#cv.weight1 = mean(residuals(weightglm1)^2/(1-hatvalues(weightglm1))^2)
#cv.weight1 = mean(residuals(weightglm1)^2/(1-hatvalues(weightglm1))^2)
#cv.weight1 = mean(residuals(weightglm1)^2/(1-hatvalues(weightglm1))^2)
#cv.weight1 = mean(residuals(weightglm1)^2/(1-hatvalues(weightglm1))^2)

#data.frame(cv.weightg1,cv.weightg2,cv.weightg3,cv.weightg4)

#favorite.table <- summary(final_model)$coefficients
#kable(favorite.table, digits=5)
```

**If our specific aims are testing whether shift and total metabolic minutes have an effect on weight gain, shouldn't the be in all the models**

Multivariate Linear Model - only GAINED
```{r}
#length(pracdat.mod2$lbs_gained)
#length(pracdat.mod2$shift)
multi.mod1 = glm(lbs_gained ~ shift, data = pracdat.mod2)


cv.weight1 = mean(residuals(multi.mod1)^2/(1-hatvalues(multi.mod1))^2)

data.frame(cv.weight1)
```


